---
title: "Tidy Tuesday (2025-10-07): EuroLeague Basketball"
format: html
date: last-modified
execute:
    cache: true
code-fold: true
---

# Intro

## Libraries

```{r}
#| label: libraries
#| echo: false

library(tidyverse)
library(scales)
library(RColorBrewer)
library(sf)
library(spData)
library(maps)
library(ggrepel)
library(colorspace)
library(leaflet)
```

## Background

Below is copied from [GitHub](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-10-07/readme.md).

This week we’re exploring EuroLeague Basketball, the premier men’s club basketball competition in Europe.

The dataset contains information on EuroLeague teams, including their country, home city, arena, seating capacity, and historical performance (Final Four appearances and titles won).

The dataset is curated from publicly available sources such as Wikipedia and official EuroLeague records, and was packaged in the EuroleagueBasketball R package, with documentation available at natanast.github.io/EuroleagueBasketball.

"The EuroLeague is the top-tier European professional basketball club competition, widely regarded as the most prestigious competition in European basketball." — EuroLeague

**Questions**

 - Which countries are most represented in the EuroLeague?
 - How do arena capacities compare across teams and countries? In R, the readr::parse_number() function might be helpful here.
 - Which clubs have been the most successful historically?

Thank you to Natasa Anastasiadou for curating this week's dataset.

## Load Data

```{r}
#| label: load-data
#| output: false

tuesdata <- tidytuesdayR::tt_load('2025-10-07')

euroleague_basketball <- tuesdata$euroleague_basketball |>
  rename_with(str_to_lower) |>
  rename_with(\(col_name) str_replace(col_name, " ", "_"))
```

## Wrangle

```{r}
#| label: wrangle-capacity

tb_capacity <- euroleague_basketball |>
  select(team, home_city, arena, capacity, country) |>
  separate_longer_delim(c(arena, capacity), ", ") |>
  mutate(capacity = as.integer(str_remove(capacity, ",")))
```


```{r}
#| label: wrangle-finalfour

tb_finalfour <- euroleague_basketball |>
  select(
    team,
    home_city,
    last_season,
    country,
    years_of_finalfour_appearances,
    years_of_titles_won
  ) |>
  separate_longer_delim(years_of_finalfour_appearances, ", ") |>
  separate_wider_delim(
    years_of_titles_won,
    ", ",
    names_sep = "",
    too_few = "align_start"
  ) |>
  mutate(
    won_title = if_any(
      contains("years_of_titles_won"),
      \(year) year == years_of_finalfour_appearances
    )
  ) |>
  mutate(won_title = replace_na(won_title, F)) |>
  select(-contains("years_of_titles_won")) |>
  rename(final_four_appearance = years_of_finalfour_appearances)
```

# Exploratory Analysis

## Most Represented Countries

```{r}
#| label: fig-most-rep

euroleague_basketball |>
  group_by(country) |>
  count() |>
  arrange(desc(n)) |>
  ungroup() |>
  mutate(country = fct_inorder(country)) |>
  ggplot(aes(country, n, fill = country)) +
  geom_col() +
  labs(
    title = "Spain has more basketball teams in the EuroLeague than any other country.",
    y = "number of teams"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Arena Capacities

```{r}
#| label: area-caps

tb_capacity_summary <- tb_capacity |>
  distinct(arena, .keep_all = T) |>
  group_by(country) |>
  summarize(country_cap = sum(capacity)) |>
  arrange(desc(country_cap)) |>
  mutate(country = fct_inorder(country))

tb_capacity_factor <- tb_capacity |>
  mutate(country = factor(country, levels = tb_capacity_summary$country)) |>
  mutate(
    team = fct_reorder(
      team,
      capacity,
      sum,
      .desc = T
    )
  ) |>
  arrange(country, team, desc(capacity)) |>
  mutate(arena = fct_inorder(arena))

tb_capacity_color <- tb_capacity_factor |>
  nest(.by = country) |>
  mutate(h = seq(0, 0.9, length.out = n_distinct(country))) |>
  unnest(data) |>
  group_by(country) |>
  mutate(
    s = seq(1, by = -0.1, length.out = length(country)),
    v = seq(1, by = -0.2, length.out = length(country)),
    hex_col = hsv(h, s, v)
  ) |>
  mutate(arena_city = str_c(arena, " (", home_city, ")"))

tb_capacity_color |>
  distinct(arena_city, .keep_all = T) |>
  ggplot(aes(country, capacity, fill = arena)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = tb_capacity_color$hex_col) +
  labs(
    title = "Arena capacities of EuroLeague Basketball teams",
    y = "arean capacity"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

tb_capacity_color |>
  distinct(arena_city, .keep_all = T) |>
  ggplot(aes(country, capacity, fill = capacity)) +
  geom_bar(position = "stack", stat = "identity") +
  viridis::scale_fill_viridis(option = "A") +
  labs(
    title = "Arena capacities of EuroLeague Basketball teams",
    y = "arean capacity"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: area-caps-databot

# all of this cell's contents generated by Databot (Claude Sonnet 4.5)

country_colors <- c(
  "Spain" = "#D32F2F", # Red
  "France" = "#1976D2", # Blue
  "Greece" = "#388E3C", # Green
  "Italy" = "#F57C00", # Orange
  "Turkey" = "#7B1FA2", # Purple
  "Israel" = "#C2185B", # Pink
  "Serbia" = "#0097A7", # Cyan
  "United Arab Emirates" = "#FBC02D", # Yellow
  "Lithuania" = "#5D4037", # Brown
  "Germany" = "#455A64", # Blue Grey
  "Monaco" = "#E64A19" # Deep Orange
)

# Remove duplicate arenas (keep one instance per unique arena)
tb_capacity_unique <- tb_capacity_factor |>
  distinct(arena, .keep_all = TRUE) |>
  arrange(country, desc(capacity))

# Create color palette with arenas sorted by capacity within each country
arena_colors <- tb_capacity_unique |>
  group_by(country) |>
  mutate(
    n_arenas = n(),
    arena_num = row_number(),
    # Create lighter/darker shades for multiple arenas in same country
    lightness = if_else(
      n_arenas == 1,
      0,
      (arena_num - 1) / (n_arenas - 1) * 0.4 - 0.2
    )
  ) |>
  ungroup() |>
  mutate(
    base_color = country_colors[as.character(country)],
    arena_color = lighten(base_color, amount = lightness)
  )

# Create named vector for the scale
arena_color_vec <- setNames(arena_colors$arena_color, arena_colors$arena)

# Reorder the arena factor to match the sorted order
tb_capacity_unique <- tb_capacity_unique |>
  mutate(arena = fct_inorder(arena))

# Create the plot with unique arenas
ggplot(tb_capacity_unique, aes(x = country, y = capacity, fill = arena)) +
  geom_col() +
  scale_fill_manual(values = arena_color_vec) +
  labs(
    title = "EuroLeague Arena Capacity by Country",
    subtitle = "Arenas sorted by capacity within each country (unique arenas only)",
    x = "Country",
    y = "Total Capacity",
    fill = "Arena"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  scale_y_continuous(labels = scales::comma)
```

This analysis is made a little bit complicated by the fact that two different Serbian teams share the same arena. 

### Experiments with Color

Did some manually experimenting before using Databot, which showed me *lighten()* from the **colorspace** package.

```{r}
#| label: colors

tb_colors <- tibble(
  h = seq(0, 0.9, length.out = 11),
  s = h,
  v = h
) |>
  expand(h, s, v) |>
  mutate(hex = hsv(h, s, v))
```

```{r}
#| label: test

h <- 0.01
s <- 0
v <- 0.1

c_col <- c(
  hsv(0.51, 1, 1),
  hsv(0.5 - h, 1 - s, 1 - v),
  hsv(0.5 - 2 * h, 1 - 2 * s, 1 - 2 * v)
)

show_col(c_col)
```

### Mapping

```{r}
#| label: maps

data(world)

world_subset <- world |>
  inner_join(
    tb_capacity_factor,
    join_by(name_long == country),
    relationship = "many-to-many"
  )

tb_capacity_gps <- world.cities |>
  mutate(
    name = case_match(
      name,
      "Tel Aviv-Yafo" ~ "Tel Aviv",
      "Vitoria" ~ "Vitoria-Gasteiz",
      "Piraieus" ~ "Piraeus",
      "Fontvieille" ~ "Monaco",
      .default = name
    )
  ) |>
  rename(home_city = name, country = country.etc) |>
  inner_join(tb_capacity_factor, join_by(home_city, country))

ggplot() +
  geom_sf(data = world) +
  geom_sf(data = world_subset, aes(fill = name_long)) +
  geom_point(
    data = tb_capacity_gps,
    aes(long, lat, color = capacity, size = capacity)
  ) +
  viridis::scale_color_viridis(option = "A") +
  geom_text_repel(
    data = distinct(tb_capacity_gps, arena, .keep_all = T),
    aes(long, lat, label = arena),
    force = 2
  ) +
  coord_sf(xlim = c(-30, 60), ylim = c(25, 75))
```

```{r}
#| label: interactive-map

# Get unique arenas
tb_capacity_gps_unique <- tb_capacity_gps |>
  distinct(arena, .keep_all = TRUE) |>
  mutate(country = factor(country, levels(tb_capacity_unique$country)))

# Create color palette for countries
pal <- colorFactor(
  palette = country_colors,
  domain = tb_capacity_gps_unique$country
)

# Create the interactive map
leaflet(tb_capacity_gps_unique) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~ sqrt(capacity) / 20, # Scale circle size by capacity
    color = ~ pal(country),
    fillColor = ~ pal(country),
    fillOpacity = 0.6,
    weight = 2,
    popup = ~ paste0(
      "<strong>",
      arena,
      "</strong><br/>",
      "Team: ",
      team,
      "<br/>",
      "City: ",
      home_city,
      "<br/>",
      "Country: ",
      country,
      "<br/>",
      "Capacity: ",
      format(capacity, big.mark = ",")
    )
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~country,
    title = "Country",
    opacity = 0.7
  )
```

Note that Monaco is a soverign city-state and microstate bording France.

## Most Winningest Teams

